---
title: "R explained in 1h45 : Exercices"
format: html
editor: visual
---

## Reminders

### What is R ?

R is a **programming language** developed to :

- Manage data : import, transform, clean, process, export ...\
- Compute **statistical analyses**, from basic to complex ones : descriptive, exploration, modelling, machine learning ...\
- Create nice **visualisations**, reports ...

[[R is free, cross-platform, open-source & collaborative !]{.smallcaps}]{style="color: #74C476"}

### How to use

You can use it directly in command line using `R` to start a session or `Rscript` to launch a full script at once. Scripts are saved with `.R` extension file.\
For reporting, statistical analyses ... it's advised to use notebooks documents, like this one, to allow you mix notes & formatting along with your code. You can use quarto `.qmd` files or Rmarkdown `.Rmd` files.

Using Rstudio allows you to navigate through files, packages and all variables stored in your environment.

## Basic operations

```{r, eval=F}
2 + 3

4 * 5

6 / 4

1,2

1.2
```

## Objects & variables

R manipulates several kind of **objects**, that can be stored in **variables** with given names, so that they can be used later\
[[Be careful to **never** use those as variable names :]{.smallcaps}]{style="color : #E39E4F;"} TRUE, FALSE, T, F, c, t, pi, data, LETTERS, letters, ...

### Numerical values

```{r}
a <- 2    # create a variable named "a" and assign its value to 2
print(a)  # displays a
a == 2    # Test if a is equal to 2  
```

Now, create a second variable named `b` with a value of 3. Test if it is equal to a.

```{r}

```

Store the result of a + b in a new variable. What happens if we change a's value to 7 ?

```{r}

```

### Vectors

Several values can be stored in a vector :

```{r}
measures <-  c()    # One vector containing 1, 10 and the values contained in a and b
integers <-  1:10		   # One vector with all integer values from 1 to 10
ducks <-  c("riri", "fifi", "loulou")    # A vector containing character strings
```

Does this work ?

```{r, eval=F}
test <-  c("riri", 1, 4:6)
```

We can compute mathematical operations on vectors :

```{r, eval=F}
measures + a 		# sum of a vector and a numerical value
integers / 2		# dividing a whole vector
```

Does this work ?

```{r, eval=F}
ducks / 2		
```

`[ ]` can be used to access elements of a vector, using *indexes* :

```{r}
integers[8]		
```

Can you change the first value of `measures` to *12* ?

```{r}

```

### Lists

Lists are another way to store values : they can be created using `list()` function :

```{r}
empty <- list()  # Create an empty list
groceries <- list("banana"=2,"apple"=5,"kiwi"=8) # a list of fruits
unnamed <- list(2,5,8) # a list of unnamed numbers
vrac <- list("A"=1:4,"B"=c(0,1),"C"="hey") # a list with multiple class of objects !
```

- They contain several objects called `elements`
- Those elements can have different classes of objects : numerical values, vectors, characters, other lists ...
- Elements are a combination of a `name`, called a key, and a `value`
- Names can be empty

To access an `element` of a list, this time use `[[ ]]` and either its index (position of the element in the list) or its name :

```{r}
groceries[["banana"]]     # Access the value stored under banana name  
groceries[[1]]            # Access the value of the first grocerires element (banana)
```

If an element is already named, you can use `$` to access it directly :

```{r}
groceries$banana          # Access the value stored under banana name 
```

Try to access an element of the list `unnamed` with `$` : what happens ?

```{r, eval=F}

```

[[TIP:]{.smallcaps}]{style="color: #1155cc"} R auto completion will help you select the right element

Use `[[ ]]` to add a named element to the list `empty`

```{r}

```

Use `[[ ]]` to add an unnamed element in the 4th position of the list `empty`

```{r}

```

What does this list contain now ?

```{r}

```

Using `c()`, we can add new values to a vector :

```{r}
measures <- c(measures, 9, 8)
```

Some values can be empty : R will use `NA` as their value

```{r}
measures[4] <- NA
measures
```

[[warning :]{.smallcaps}]{style="color: #E39E4F"} Objects with missing values might raise unexpected behaviors

### Functions

A function has :

- A `name`, to execute the function or look for help\
- One or more `arguments` : some are necessary for the function to run, others have default values
- An `output` : the result of the computations on the given arguments. With complex functions, this will often be a list

Look at the function `mean` help :

```{r}
?mean
```

We use `( )` to give the function its `arguments` before running : for mean, a vector of numerical values. How can you compute the mean of measures ?

```{r}

```

If any of the values is missing, `mean` will return a missing value `NA`. We can use the argument `na.rm` to ignore them :

```{r}

```

Arguments can be :\
- named =\> they can be written in any order\
- unnamed =\> R will expect them in their appearing order

We can store the output of a function in a variable as well :

```{r}

```

**Variable != Arguments**\
`mean_measures` is now stored in the environment, but `x` and `na.rm` are not : they where only given to the mean function

Finally, you can create your own functions by defining its name, arguments, output and the list of operations needed to compute the results you want

```{r}
square <- function(x) {
  # as many lines as you want here !
  output = x * x
  return(output)
}
```

This can be useful if you need to repeat the same computations several times, or if you need to modify some existing functions

## Memos for later

### Manipulate basic objects

| Symbol | Goal | Examples |
|------------------------|------------------------|------------------------|
| `=` or `<-` or `->` | assign a `value` to a `variable` or an `argument` | a `<-` 10 or read.table(file `=` "file.txt") |
| `( )` | to give `arguments` to a function | mean`(`x = c`(`10,6,8,2`))` |
| `[ ]` | get a value from a vector | measures`[`1`]` or measures`[`2:8`]` or measures`[`c(3,6,4)`]` |
| `[[ ]]` or `$` | get an element from a list | groceries`$`banana or groceries`[[`"banana"`]]` or groceries`[[`1`]]` |
| `[ , ]` ou `$` | extract elements from a data-frame | annotation`$`geneID or annotation`[`1:5`,`geneID`]` |

### Good practices

| To | Do this | Avoid this |
|------------------------|------------------------|------------------------|
| Clean your environment | `rm(list = ls())` | mixing-up variables |
| Use explicit names | expression_data, annotation, groceries_list | dat, var1, dfslkgj |
| Use comments | with `#`, using lines, stars ... | obscure functions, 100+ lines without comment |
|  | use spaces, skip lines, add blocks ... | all commands in one block |

[Tidyverse guidelines](https://style.tidyverse.org/syntax.html){.external target="_blank"}

### Logical operators

| Operator | Description      | Example        | Result  |
|----------|------------------|----------------|---------|
| `==`     | Equal to         | `5 == 3`       | `FALSE` |
| `!=`     | Not equal to     | `5 != 3`       | `TRUE`  |
| `>`      | Greater than     | `5 > 3`        | `TRUE`  |
| `<`      | Less than        | `5 < 3`        | `FALSE` |
| `>=`     | Greater or equal | `5 >= 5`       | `TRUE`  |
| `<=`     | Less or equal    | `5 <= 3`       | `FALSE` |
| `&`      | Element-wise AND | `TRUE & FALSE` | `FALSE` |
| `|`      | Element-wise OR  | `TRUE | FALSE` | `TRUE`  |
| `!`      | NOT              | `!TRUE`        | `FALSE` |

### Aritmetic operators

| Operator | Description      | Example   | Result |
|----------|------------------|-----------|--------|
| `+`      | Addition         | `5 + 3`   | `8`    |
| `-`      | Subtraction      | `5 - 3`   | `2`    |
| `*`      | Multiplication   | `5 * 3`   | `15`   |
| `/`      | Division         | `5 / 2`   | `2.5`  |
| `^`      | Exponentiation   | `2 ^ 3`   | `8`    |
| `%%`     | Modulus          | `5 %% 2`  | `1`    |
| `%/%`    | Integer Division | `5 %/% 2` | `2`    |
